\mbox{Specifying Imperative Higher Order Programs}

### 2009-09-30 papers/esop2010/main.tex
go

\newcommand{\wpre}{ {\bf wp}}
\newcommand{\sep}{ ~|~ }
\newcommand{\letml}{{\bf let}}
\newcommand{\inml}{{\bf in}}
\newcommand{\ifml}{{\bf if}}
\newcommand{\thenml}{{\bf then}}
\newcommand{\elseml}{{\bf else}}
\newcommand{\refml}{{\bf ref}}
\newcommand{\propml}{{\bf prop}}
\newcommand{\boolml}{{\bf bool}}
\newcommand{\recml}{{\bf rec}}
\newcommand{\efft}[1]{ \langle #1 \rangle }
\newcommand{\alist}[1]{ \overline{#1} }

@@

\begin{figure}[htpb]
\begin{mathpar}
  { \inferrule*[Left=Var] {Γ(x) = ∀\alist{χ}.τ} {Γ|-_v x~\alist{κ} :
  τ[\alist{χ}|->\alist{κ}]} } \and
  { \inferrule*[Left=PureFun] {Γ,x : τ' |-_v t : τ} {Γ|-_v λx : τ'.t : τ' ->
  τ} } \and
  { \inferrule*[Left=Rec] 
  { Γ' = Γ,~f : τ' ->^ε τ,~ x : τ' \\ Γ' |- t : τ,ε \\ Γ', cur : \efft{ε} |- p : \propml \\ 
  Γ', old, cur : \efft{ε} |- q : τ -> \propml } 
      {Γ|-_v \recml~f~(x : τ').\{p\}t\{q\} : τ' ->^ε τ} 
  } \and
  { \inferrule*[Left=PureApp]
    {Γ |-_v : t1 : τ' -> τ \\ Γ|-_v : t2 : τ'}
    {Γ |-_v : t1 ~ t2 : τ}
  } \and
  { \inferrule*[Left=App]
    {Γ |-_v : t1 : τ' ->^ε τ \\ Γ|-_v : t2 : τ'}
    {Γ |- : t1 ~ t2 : τ, ε}
  } \and
  { \inferrule*[Left=ite]
    { Γ |- _v : t1 : \boolml \\ Γ |- t2 : τ, ε1 \\ Γ |-t3 : τ, ε2 }
    { Γ |- \ifml~ t1 ~\thenml~ t2 ~\elseml ~ t3 : τ, ε1  ε2 }
  } \and
  { \inferrule*[Left=Letv]
    { Γ, \alist{χ} |-_v : t1 : τ' \\ Γ, x : ∀\alist{χ}.τ' |- t2 : τ,ε }
    { Γ |- \letml~ x~[\alist{χ}] = t1 ~ \inml~ t2 : τ, ε}
  } \and
  { \inferrule*[Left=Let]
    { Γ |- : t1 : τ', ε1 \\ Γ, x : τ' |- t2 : τ, ε2 }
    { Γ |- \letml~ x = t1 ~ \inml~ t2 : τ, ε1  ε2 }
  }
\end{mathpar}
  \caption{The Typing relation}
  \label{fig:typing}
\end{figure}

Again in order to simplify future definitions, the definition typing relation
differs somewhat from the standard way (see Fig.\ref{fig:typing}). First of
all, we present the language in {\em A-normal form} (ANF), where all values
obtained by intermediate computations are named. However, our definition of
values is not syntactical: our ``semantic'' values are objects which are
either syntactic values (abstractions and variables) or are obtained by {\em
pure application}, the application which involves the pure arrow $τ->τ$.
Furthermore, the system is a type and effect system. There are in fact two
typing relations, the first one is of the form $Γ|-_v : t : τ$. In this case,
$t$ is a semantic value of type $τ$ and cannot have any effect. In the other
case, we have $Γ|- : t : τ, ε$, where $t$ is an expression of type $τ$ and
effect $ε$.

Effects can be produced only by function application, where the impure arrow
$τ->^ετ$ is involved.

predefined : \propml \boolml

unusual : 
* ANF 
* values defined by types
* two different types of arrows
* distinction bool and prop
* notion of substitution - correspondence χ κ
* TODO logic typing

missing : 
* value injection
* subeffecting

>>>>

\newcommand{\wpre}{ {\bf wp}}
\newcommand{\sep}{ ~|~ }
\newcommand{\letml}{{\bf let}}
\newcommand{\inml}{{\bf in}}
\newcommand{\ifml}{{\bf if}}
\newcommand{\thenml}{{\bf then}}
\newcommand{\elseml}{{\bf else}}
\newcommand{\refml}{{\bf ref}}
\newcommand{\propml}{{\bf prop}}
\newcommand{\boolml}{{\bf bool}}
\newcommand{\recml}{{\bf rec}}
\newcommand{\efft}[1]{ \langle #1 \rangle }
\newcommand{\lift}[1]{ \langle #1 \rangle }
\newcommand{\alist}[1]{ \overline{#1} }
\newcommand{\Trueml}{{\bf True}}
\newcommand{\Falseml}{{\bf False}}
\newcommand{\trueml}{{\bf true}}
\newcommand{\falseml}{{\bf false}}
\newcommand{\coq}{Coq}

\newcommand{\ceil}[1]{\lceil #1 \rceil}

@@

\begin{figure}[htpb]
\begin{mathpar}
  { \inferrule*[Left=Var] {Γ(x) = ∀\alist{χ}.τ} {Γ|-_v x~\alist{κ} :
  τ[\alist{χ}|->\alist{κ}]} } \and
  { \inferrule*[Left=PureFun] {Γ,x : τ' |-_v t : τ} {Γ|-_v λx : τ'.t : τ' ->
  τ} } \and
  { \inferrule*[Left=Rec] 
  { Γ' = Γ,~f : τ' ->^ε τ,~ x : τ' \\ Γ' |- t : τ,ε \\ Γ', cur : \efft{ε} |-_l p : \propml \\ 
  Γ', old, cur : \efft{ε} |-_l q : τ -> \propml } 
      {Γ|-_v \recml~f~(x : τ').\{p\}t\{q\} : τ' ->^ε τ} 
  } \and
  { \inferrule*[Left=PureApp]
    {Γ |-_v t1 : τ' -> τ \\ Γ|-_v t2 : τ'}
    {Γ |-_v  t1 ~ t2 : τ}
  } \and
  { \inferrule*[Left=App]
    {Γ |-_v t1 : τ' ->^ε τ \\ Γ|-_v t2 : τ'}
    {Γ |- t1 ~ t2 : τ, ε}
  } \and
  { \inferrule*[Left=ite]
    { Γ |- _v t1 : \boolml \\ Γ |- t2 : τ, ε1 \\ Γ |- t3 : τ, ε2 }
    { Γ |- \ifml~ t1 ~\thenml~ t2 ~\elseml ~ t3 : τ, ε1  ε2 }
  } \and
  { \inferrule*[Left=Letv]
    { Γ, \alist{χ} |-_v t1 : τ' \\ Γ, x : ∀\alist{χ}.τ' |- t2 : τ,ε }
    { Γ |- \letml~ x~[\alist{χ}] = t1 ~ \inml~ t2 : τ, ε}
  } \and
  { \inferrule*[Left=Let]
    { Γ |- t1 : τ', ε1 \\ Γ, x : τ' |- t2 : τ, ε2 }
    { Γ |- \letml~ x = t1 ~ \inml~ t2 : τ, ε1  ε2 }
  } \and
  { \inferrule*[Left=Value]
  { Γ |-_v t : τ  }
  { Γ |- t : τ,\emptyset  }
  } \and
  { \inferrule*[Left=Logic]
    { Γ |-_v t : τ  }
    { Γ |-_l t : \ceil{τ} }
  }
\end{mathpar}
  \caption{The Typing relation}
  \label{fig:typing}
\end{figure}

\begin{figure}[htpb]
  \begin{eqnarray*}
    \ceil{α} &=& α\\
    \ceil{τ\ref_ρ} &=& \ceil{τ}\ref_ρ\\
    \ceil{τ -> τ'} &=& \ceil{τ} -> \ceil{τ'}\\
    \ceil{τ \times τ'} &=& \ceil{τ} \times-> \ceil{τ'}\\
    \ceil{\efft{ε}} &=& \efft{ε}\\
    \ceil{τ ->^ε τ'} &=& (\ceil{τ} -> \efft{ε} -> \propml)\times (\ceil{τ} ->
    \efft{ε} -> \efft{ε} -> \ceil{τ'} -> \propml)()
  \end{eqnarray*}
  \caption{Lifting types}
  \label{fig:typelift}
\end{figure}

Again in order to simplify future definitions, the definition typing relation
differs somewhat from the standard way (see Fig.\ref{fig:typing}). First of
all, we present the language in {\em A-normal form} (ANF), where all values
obtained by intermediate computations are named. However, our definition of
values is not syntactical: our ``semantic'' values are objects which are
either syntactic values (abstractions and variables) or are obtained by {\em
pure application}, the application which involves the pure arrow $τ->τ$.
Furthermore, the system is a type and effect system. There are in fact two
typing relations, the first one is of the form $Γ|-_v  t : τ$. In this case,
$t$ is a semantic value of type $τ$ and cannot have any effect. It can be seen
that the set of terms described by $|-_v$ is the precisely the one obtained by
combining syntactic values and pure application in a well-typed way.

In the other case, we have $Γ|- t : τ, ε$, where $t$ is an expression of
type $τ$ and effect $ε$. Impure applications, \letml-expressions and
\ifml-expressions can only be typed in this way. Effects can be produced only
by function application, where the impure arrow $τ'->^ετ$ is involved. In this
case, the effect of the application is precisely the effect $ε$ of the arrow,
because the two subterms are semantic values and don't have any effect.

The typing relies on two predefined types, \propml\ and \boolml. \propml\ is
the logical truth and contains the values \Trueml\ and \Falseml, while
\boolml\ describes boolean (decidable) properties and contains \trueml\ and
\falseml. The distinction between \propml\ and \boolml\ is the same one as in
\coq, for example. Properties expressed in \propml\ are potentially
undecidable and one cannot take decisions by testing their truth value. On the
other hand, \boolml\ talks about decidable properties and so it makes sense to
expect the predicate of the \ifml-statement to be of type \boolml.

unusual : 
* ANF 
* values defined by types
* two different types of arrows
* distinction bool and prop
* notion of substitution - correspondence χ κ

missing : 
* value injection
* subeffecting
* logic typing

%TODO is the LOGIC rule correct ??
